---
title: "Data Manipulation with dplyr"
author: "Jingyi"
date: "9/23/2019"
output: github_document
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load the library
library(tidyverse)

# what is this?
options(tibble.print_min = 3)

# load datasets, parse columns and clean data names
litters_data = read_csv("./data/FAS_litters.csv",
  col_types = "ccddiiii")
litters_data = janitor::clean_names(litters_data)

# If we look at litters_data, there is a variable called group.
# R doesn't know group, but R knows about litters_data.
# Tell R where to look, and then you can get the data inside the datafram under specific variable

pull(litters_data, group)

pups_data = read_csv("./data/FAS_pups.csv",
  col_types = "ciiiii")
pups_data = janitor::clean_names(pups_data)

```


##Selecting

For a given analysis, you may only need a subset of the columns in a data table; extracting only what you need can helpfully de-clutter, especially when you have large datasets.

Select columns using *select*.

**select(dataname, variable*)**


```{r select}

# You can specify the columns you want to keep by naming all of them:

select(litters_data, group, litter_number, gd0_weight, pups_born_alive)
# return with the order above

# If you want to select everything, here is a helper function:
select(litters_data, litter_number, group, everything())

# If you want to select column names that start with “pups”, here is a helper function:
select(litters_data, litter_number, gd0_weight, starts_with("pups"))


# You can specify the specify a range of columns to keep:
select(litters_data, group:gd_of_birth)

# You can also specify columns you’d like to remove:

select(litters_data, -pups_survive)

# You can rename variables as part of this process:

select(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)

# If all you want to do is rename something, you can use rename instead of select. This will rename the variables you care about, and keep everything else:

rename(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)

```

There are some handy helper functions for select; read about all of them using ?select_helpers. The one I use most frequently is everything(), which is handy for reorganizing columns without discarding anything:

```{r helper function}

select(litters_data, litter_number, group, everything())

```


In larger datasets, I also use starts_with(), ends_with(), and contains() often.

Lastly, like other functions in dplyr, select will export a dataframe even if you only select one column. Mostly this is fine, but sometimes you want the vector stored in the column. 

To pull a single variable, use pull.


###Leanring Assessment#1:
In the pups data, select the columns containing litter number, sex, and PD ears.

```{r Learning Assessment #1}

select(pups_data, litter_number:pd_ears)

```

##Fliter


Some data tables will include rows you don’t need for your current analysis.


Although you could remove specific row numbers using base R, you shouldn’t – this might break if the raw data are updated, and the thought process isn’t transparent. 

Instead, you should filter rows based on logical expressions using the filter function. 

Like select, the first argument to filter is the dataframe you’re filtering; all subsequent arguments are logical expressions.


You will often filter using comparison operators (>, >=, <, <=, ==, and !=). 


You may also use %in% to detect if values appear in a set, and is.na() to find missing values.


The results of comparisons are logical – the statement is TRUE or FALSE depending on the values you compare – and can be combined with other comparisons using the logical operators & and |, or negated using !.


Some ways you might filter the litters data are:

* gd_of_birth == 20
* pups_born_alive >= 2
* pups_survive != 4
* !(pups_survive == 4)
* group %in% c("Con7", "Con8")
* group == "Con7" & gd_of_birth == 20


A very common filtering step requires you to omit missing observations. You can do this with filter, but I recommend using drop_na from the tidyr package:

* drop_na(litters_data) will remove any row with a missing value.
* drop_na(litters_data, wt_increase) will remove rows for which wt_increase is missing.


Filtering can be helpful for limiting a dataset to only those observations needed for an analysis. 


However, I recommend against the creation of many data subsets (e.g. one for each group). This can clutter up your workspace, and we’ll see good tools for the analysis of subsets before long.

#Learning Assessment #2:

In the pups data:
* Filter to include only pups with sex 1
* Filter to include only pups with PD walk less than 11 and sex 2

```{r Leanring Assessment #2}

filter(pups_data, sex == 1)

filter(pups_data, sex == 2 & pd_walk < 11)

```

## Mutate

Sometimes you need to select columns; sometimes you need to change them or create new ones. You can do this using mutate.

The example below creates a new variable measuring the difference between gd18_weight and gd0_weight and modifies the existing group variable.

```{r}

mutate(litters_data, wt_gain = gd18_weight - gd0_weight, group = str_to_lower(group))
# Question: What does this str_to_lower(group) do?
# make all string variables to their lower cases
# I can modify and overwrite the existing variables.

```

A few things in this example are worth noting:

* Your new variables can be functions of old variables
* New variables appear at the end of the dataset in the order that they are created
* You can overwrite old variables
* You can create a new variable and immediately refer to (or change) it


Creating a new variable that does exactly what you need can be a challenge; the more functions you know about, the easier this gets.


##Learning Assessment #3


In the pups data:

* Create a variable that subtracts 7 from PD pivot

* Create a variable that is the sum of all the PD variable

```{r}

mutate(pups_data, pd_pivotsub7 = pd_pivot - 7, pd_sum = pd_ears + pd_eyes + pd_pivot + pd_walk)

```


## arrange


In comparison to the preceding, arranging is pretty straightforward. 


You can arrange the rows in your data according to the values in one or more columns:

```{r arrange}

head(arrange(litters_data, group, pups_born_alive), 10)

```

## %>%


We’ve seen several commands you will use regularly for data manipulation and cleaning.

You will rarely use them in isolation.

For example, suppose you want to load the data, clean the column names, remove pups_survive, and create wt_gain.

There are a couple of options for this kind of multi-step data manipulation:

* define intermediate datasets (or overwrite data at each stage)
* nest function calls

The following is an example of the first option:

```{r}

# load the data
litters_data_raw = read_csv("./data/FAS_litters.csv",
  col_types = "ccddiiii")

# clean data names
litters_data_clean_names = janitor::clean_names(litters_data_raw)

# remove pups_survive
litters_data_selected_cols = select(litters_data_clean_names, -pups_survive)

# create wt_gain
litters_data_with_vars = 
  mutate(
    litters_data_selected_cols, 
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group))

# get rid of all rows with missing value 
litters_data_with_vars_without_missing = 
  drop_na(litters_data_with_vars, wt_gain)

# show the data set
litters_data_with_vars_without_missing

```

Below, we try the second option:

```{r}

litters_data_clean = 
  drop_na(
    mutate(
      select(
        janitor::clean_names(
          read_csv("./data/FAS_litters.csv", col_types = "ccddiiii")          # load data
          ),        # clean the names
      -pups_survive
      ),      # remove pups_survive
    wt_gain = gd18_weight - gd0_weight,   # create wt_gain value
    group = str_to_lower(group) # make everything string variables their lower cases
    ),
  wt_gain     # create wt_gain variable and add this into the dataframe
  )

litters_data_clean # view the dataset

```


These are both confusing and bad: the first gets confusing and clutters our workspace, and the second has to be read inside out.


Piping solves this problem. 


It allows you to turn the nested approach into a sequential chain by passing the result of one function call as an argument to the next function call:

```{r}

# same thing with pipe

litters_data = 
  read_csv("./data/FAS_litters.csv", col_types = "ccddiiii") %>%                 # load the dataset and then
  janitor::clean_names() %>%    # clean the names and then
  select(-pups_survive) %>%     # remove pups_survive and then
  mutate(
    wt_gain = gd18_weight - gd0_weight,  # create wt_gain variable and
    group = str_to_lower(group)) %>%    # make group variables into their lower cases and then
  drop_na(wt_gain)              # remove rows with missing wt_gain values

litters_data   # veiw dataset

```


All three approaches result in the same dataset, but the piped commands are by far the most straightforward. 


The easiest way to read %>% is “then”; the keyboard shortcuts are Ctrl + Shift + M (Windows).


The functions in dplyr (and much of the tidyverse) are designed to work smoothly with the pipe operator. 

By default, the pipe will take the result of one function call and use that as the first argument of the next function call; by design, functions in dplyr will take a tibble as an input and return a tibble as a result. 


As a consequence, functions in dplyr are easy to connect in a data cleaning chain.


You can make this more explicit by using . as a placeholder for the result of the preceding call:


```{r using . as a placeholder for result of the preceding call}

litters_data = 
  read_csv("./data/FAS_litters.csv", col_types = "ccddiiii") %>% # load dataset
  janitor::clean_names(dat = .) %>% # clean names
  select(.data = ., -pups_survive) %>% # remove pups_survive
  mutate(.data = ., # preceding dataset
    wt_gain = gd18_weight - gd0_weight, # add wt_gain variable
    group = str_to_lower(group)) %>%  # make them lower cases
  drop_na(data = ., wt_gain) # drop missing wt_gain value rows

```


In this example, the dataset argument is called dat in janitor::clean_names, .data in the dplyr functions, and data in drop_na – which is definitely confusing.


In the majority of cases (and everywhere in the tidyverse) you’ll elide the first argument and be happy with life, but there are some cases where the placeholder is necessary.


For example, to regress wt_gain on pups_born_alive, you might use:

```{r}

litters_data %>%
  lm(wt_gain ~ pups_born_alive, data = .) %>%
  broom::tidy()

```


There are limitations to the pipe.


You shouldn’t have sequences that are too long; there isn’t a great way to deal with multiple inputs and outputs; and (since it’s not base R) not everyone will know what %>% means or does.


That said, compared to days when R users only had the first two options, life is much better!


## Learning Assessment #4

Write a chain of commands that:

* loads the pups data
* cleans the variable names
* filters the data to include only pups with sex 1
* removes the PD ears variable
* creates a variable that indicates whether PD pivot is 7 or more days

```{r Learning Assessment #4}

pups_data = 
  read_csv("./data/FAS_pups.csv",
  col_types = "ciiiii") %>%         # load data
  janitor::clean_names(dat = .) %>% # clean names
  filter(.data = ., sex == 1) %>%   # filter data to include only pups with sex 1
  select(.data = ., -pd_ears) %>%   # remove PD ears variable
  mutate(.data = .,
         pd_pivot_gt7 = pd_pivot >= 7)             # create a variable that indicates whether PD pivot is 7 or more days

pups_data 

```

