---
title: "Data Import"
author: "Jingyi"
date: "9/22/2019"
output: github_document
---

I downloaded the data and created a GitHub repo + local R project.

**How to move the datasets to my repo/directory?**
*Is https://help.github.com/en/articles/moving-a-file-to-a-new-location correct?*
**Follow the lecture, and just add the data files into the local R folder.**

```{r setup, include = FALSE}

# What does this mean?
knitr::opts_chunk$set(echo = TRUE)

# load the library
library(tidyverse)

```

## Load in a dataset

*I can use library(here) and here::here("file_I_want.csv") to find where my file is.*

**Never ever use setwd()**

```{r}

## reads in a dataset

# Load from a relative path.
# Question: What's the difference between read_csv and read.csv?
litters_data = read_csv(file = "./data/FAS_litters.csv")


## use janitor::clean_names() to clean up variable names after importing
## This will take the column names are and convert them to lower snake case.

names(litters_data)

litters_data = janitor::clean_names(litters_data)

names(litters_data)

# Question: Why my output are not italized?

```

**The package::function syntax lets you use a function from a package without loading the whole library.**
*Figure out what is conflicted package.*

***Leanring Assessment***

Make sure you are able to load the FAS_pups.csv dataset. Use both absolute and relative paths. Quit R Studio and move the directory containing your project, data, and R Markdown document. Repeat the previous data import process; do both absolute and relative paths still work?

```{r learning assessment #1}

# load from absolute path
FAS_pups_absolute = read_csv(file = "C://Users//Jingyi//Desktop//Columbia_University//Fall2019//P8105_Data_Science_I//Lecture In-Class Codes//Data_Wrangling_I//data_wrangling_i//Data//FAS_pups.csv")

# load from relative path
FAS_pups_relative = read.csv(file = "./data/FAS_pups.csv")

```

Good Try!

##Looking at data

